---
title: "Collate GWAS Information"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

This script creates a file describing each GWAS, including information stored within the SQL database (as input manually), as well as the following additional information:

- Title, abstract, link to reference from PubMed
- LDSC statistics

***

# Set up RMD options and environment

```{r}
knitr::opts_chunk$set(echo = TRUE,
                      comment=NA,
                      prompt=FALSE,
                      cache=FALSE)
```

Clear global environment
```{r}
remove(list = ls())
```

***

# Set output parameter

```{r}
opt<-list()

# This is an example output parameter
opt$output<-'/users/k1806347/brc_scratch/Data/GWAS_sumstats/Rosalind_repository/example'

# Identify the directory for the output
opt$output_dir<-dirname(opt$output)

# Create output directory if it doesn't exist
if(!file.exists(opt$output_dir)){
  dir.create(opt$output_dir)
}
```

***

# Load required packages

```{r}

library(data.table)
library(easyPubMed)

```

***

# Load up-to-date SQL database csv

```{r}
# I don't know where the location of this csv is, so this is an example
csv_new<-fread('/users/k1806347/brc_scratch/Data/GWAS_sumstats/Rosalind_repository/download_csv_test.csv')

```

***

# Create or update GWAS inforamtion file

The process of downloading PubMed information is slow, and therefore it is most efficient to update a pre-existing GWAS information file to include new entries.

```{r}
# Read in previous GWAS information file if available
opt$gwas_info<-NA

if(!is.na(opt$gwas_info)){
  # Read previously created gwas_info file
  gwas_info<-fread(opt$gwas_info)
  
  # Identify entries in csv_new that are not in gwas_info
  new_gwas<-csv_new[!(csv_new$code %in% gwas_info$code),]
  
  # Quit if no new entries are detected
  if(sum(!(csv_new$code %in% gwas_info$code)) == 0){
    cat('No new entries detected.\n')
    q()
  } else {
    cat(dim(new_gwas)[1],'new entries detected.\n')
  }

} else {
  # No previous gwas_info file exists, so run for all entries in csv_new
  new_gwas<-csv_new
  cat(dim(new_gwas)[1],'new entries detected.\n')
}

```

***

# Download PubMed information

```{r}

new_gwas$article_title<-NA
new_gwas$article_abstract<-NA

for(i in which(!is.na(new_gwas$pmid))){
  pmid_temp<-get_pubmed_ids(new_gwas$pmid[i], api_key = NULL)
  dat<-fetch_pubmed_data(pmid_temp, format='xml')
  new_gwas$article_title[i] <- custom_grep(dat, "ArticleTitle", "char")
  new_gwas$article_abstract[i] <- paste0(custom_grep(dat, "AbstractText", "char"), collapse=' ')
  cat(paste0(round(i/sum(!is.na(new_gwas$pmid)),2)*100,"% complete\n"))
}

```

***

# Integrate LDSC statistics

```{r}

new_gwas$ldsc_h2_observed<-NA
new_gwas$ldsc_h2_se<-NA
new_gwas$ldsc_lambda_GC<-NA
new_gwas$ldsc_mean_chi2<-NA
new_gwas$ldsc_intercept<-NA
new_gwas$ldsc_intercept_se<-NA
new_gwas$ldsc_ratio<-NA
new_gwas$ldsc_ratio_se<-NA

ldsc_logs<-list.files(path='/scratch/groups/ukbiobank/sumstats/munged', pattern='_herit.log')
ldsc_logs<-gsub('_herit.log','',ldsc_logs)

for(i in 1:dim(new_gwas)[1]){
  if((new_gwas$code[i] %in% ldsc_logs)){
    ldsc_logs_i<-read.fwf(paste0('/scratch/groups/ukbiobank/sumstats/munged/',new_gwas$code[i],'_herit.log'),widths = 1000000)
        
    new_gwas$ldsc_h2_observed[i]<-as.numeric(gsub(' .*','', gsub('Total Observed scale h2: ','',ldsc_logs_i[grepl('Total Observed scale h2: ', ldsc_logs_i$V1),])))
    new_gwas$ldsc_h2_se[i]<-as.numeric(gsub("\\)",'', gsub(".*\\(",'', gsub('Total Observed scale h2: ','',ldsc_logs_i[grepl('Total Observed scale h2: ', ldsc_logs_i$V1),]))))
    new_gwas$ldsc_lambda_GC[i]<-as.numeric(gsub('Lambda GC: ','',ldsc_logs_i[grepl('Lambda GC: ', ldsc_logs_i$V1),]))
    new_gwas$ldsc_mean_chi2[i]<-as.numeric(gsub("Mean Chi\\^2: ",'',ldsc_logs_i[grepl("Mean Chi", ldsc_logs_i$V1),]))
    new_gwas$ldsc_intercept[i]<-as.numeric(gsub(' .*','', gsub("Intercept: ",'',ldsc_logs_i[grepl("Intercept: ", ldsc_logs_i$V1),])))
    new_gwas$ldsc_intercept_se[i]<-as.numeric(gsub("\\)",'', gsub(".*\\(",'', gsub("Intercept: ",'',ldsc_logs_i[grepl("Intercept: ", ldsc_logs_i$V1),]))))
    
    if((new_gwas$ldsc_intercept[i]-1)/(new_gwas$ldsc_mean_chi2[i]-1) > 0){
      new_gwas$ldsc_ratio[i]<-as.numeric(gsub(' .*','', gsub("Ratio: ",'',ldsc_logs_i[grepl("Ratio: ", ldsc_logs_i$V1),])))
      new_gwas$ldsc_ratio_se[i]<-as.numeric(gsub("\\)",'', gsub(".*\\(",'', gsub("Ratio: ",'',ldsc_logs_i[grepl("Ratio: ", ldsc_logs_i$V1),]))))
    } else {
      new_gwas$ldsc_ratio[i]<-'<0'
      new_gwas$ldsc_ratio_se[i]<-'NA'
    }
  }
}

```

***

# Write out updated GWAS information file

```{r}

# Combine with previously created gwas_info file if present
if(is.na(opt$gwas_info) == F){
  new_gwas<-rbind(gwas_info, new_gwas)
}

# Save new/updated gwas_info
write.csv(new_gwas, paste0(opt$output,'.',Sys.Date(),'.csv'), row.names=F, quote=T)

```


