---
title: "rg results summary tables template"
author: "Abigail ter Kuile"
date: "28/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment=NA,
                      prompt=FALSE,
                      cache=FALSE)
```

Clear global environment
```{r Clear global environment}
remove(list = ls())
```

```{r Packages}
library(summarytools)
library(googlesheets4)
library(tidyverse)
```

Retrieve the recent date
```{r Recent date}
date = Sys.Date()
date
```

What input do we need for this rmarkdown?
LDSC rg results between a selected trait of interest and all traits in the summary statistics repository (or a shorter list of specific list of traits). All traits have had mhc region removed before rg estimation (and are significantly heritable?). rg results are only calculated in one direction to avoid repetition. 


1) Read in list of traits to be analysed (in GWAS code format e.g. ADHD05, mhc region removed )

Option 1: up to date list of all GWAS traits for rg analysis
```{r read in up to date list of traits}
trait_list_all <- c("insert/file/path/list.txt")
```

Option 2: shorter list of specific traits for rg analysis
```{r read in sub selected trait list}
trait_list_short <- c("insert/file/path/list.txt")
```

2) Read in Up to date table with full GWAS information of each trait. Currently in csv format linked with the flask app, or in the google sheet.

# option 1: CSV
```{r csv read in full GWAS info all traits}
GWAS_info_sheet <- c("insert/file/path/sheet.csv")
```

# option 2: google sheet

Login to googlesheets
```{r Login to google sheets}
gs4_auth(email = "emailname@kcl.ac.uk")
```

```{r google read in full GWAS info all traits}
GWAS_info_sheet <- read_sheet(
  ss = "https://docs.google.com/spreadsheets/d/1Wmj4Qzumkzks6541vQJqxaUBxjfL4KLQo7_K_oquacI/edit#gid=0", #link to google doc
  sheet = "SGDP_GWASLIST_EDITTHIS", #sheet name with data in
  col_names = TRUE, #  use the first row as column names
  trim_ws = TRUE, # Trims whitespace,
  guess_max = 600, # Max number of data rows to use for guessing column types +++ATK do we need this? change to +700?
  na = c("", "NULL", "NA") #Character vector of strings to interpret as missing values. 
  )

colnames(GWAS_info_sheet)

GWAS_info_sheet
```

```{r check mhc removed }
# add once we have decided what format this is in (remove using SNP list?)
```

```{r check traits h2 significantly greater than 0}

# e.g. h2_observed+/-1.96*h2_se = upper/lower CI
# h2_observed - CI_upper 
#https://www.tidyverse.org/blog/2021/08/infer-1-0-0/
```

3) Read in rg data for your trait of interest

+++ATK: this depends on if we are using LDSC in python to run rg, or are we using Genomic SEM in R to run rg? 

if LDSC:
convert already existing bash scripts to R - or just read bash script?
```{r read in rg results}

```

```{r select rg results from log out put files}

```

2) Joining datasets:
#Step) join the rg results with the GWAS information file to get full GWAS info for each trait
Merge 
```{r join info GWAS sheet with rg results}
rgs_gwas_info <- 
  left_join(
    x = rgs,
    y = GWAS_info_sheet,
    by = c("p1" = "Code")) ##note: p1 is for LDSC-python. Change if we use LDSC-Gsem to whatever we specify for that column (phenotype?)
    
colnames(rgs_gwas_info)
```

3) Filtering
For both summary table for plotting and supplementary table:
#step) Only extract GWAS traits that are not private. 
Filter by permissions and exclude all private sumstats
NOTE: need to simplify the current permissions category 
```{r filter based on permissions}
rgs_filtered <- rgs_gwas_info %>% 
  filter(
    permissions != "PRIVATE" 
  )
```

#optional step if only interested in certain trait categories
```{r by trait cateogry}
rgs_filtered_cat <- rgs_filtered %>%
  filter(
    category = "insert_category" #add category as relevent
  )
```

#step) Filter p-values by bonferonni threshold 
Adapt Helena's script in python that already exists calculating bonferonni significance?
Create a new data frame with only bonf signif traits for summary table for paper plotting of top traits? or just add a new column with 'yes/no' for bonferonnni significant? Johan's preference: stick to one dataframe and filter in the plotting step
For supplementary table: how do we want to indicate which traits are bonferonni significant? e.g. bold text? A separate column saying 'significant/ non significant'? Colours??
```{r create bonferonni column}
rgs_filtered <- rgs_filtered %>%
  mutate(
   P.Bonferroni =
  p.adjust(
    p = p,
    method = "bonferroni" # Bonferroni correction method (p-values are multiplied by the number of comparisons)
    )
  )

summary(
  rgs_filtered$P.Bonferroni
  )
```

```{r create FDR column }
rgs_filtered <- rgs_filtered %>%
  mutate(
   FDR.Q.value =
  p.adjust(
    p = p,
    method = "BH" # Benjamini Hochberg method (the expected proportion of false discoveries amongst the rejected hypotheses)
    )
  )

summary(
  rgs_filtered$FDR.Q.value
  )
```

Add column with FDR significant label
```{r Add column with FDR significant label}
rgs_filtered <-
  rgs_filtered %>%
  mutate(
    FDR_significant =
      case_when(
        FDR.Q.value < 0.05 ~ "significant",
        FDR.Q.value > 0.05 ~ "not"
      )
  )

rgs_filtered %>%
  summarise(FDR_significant)
```

```{r convert table to matrix for matrix decomposition}
#find Helena's script for this 
```

note: we already have this script in python
```{r matrix decomposition for bonferonni threshold}
rgs_filtered <- rgs_filtered %>%
  mutate(
   P.Bonferroni_matrix =
     #find Helena's script for this?
```

Create new dataframes for supplmentary tables

#create supplementary tables
Select columns for supplementary tables for paper. Decide on columns. 
e.g. Watson et al 2019 (AN GWAS) columns: Phenotype, rg, se, P (green = bonferoni significant), Phenotype category, PMID, code, ancestry. 
but could also include Z value of rg, gcov_int, gcov_int_se, h2_obs, h2_obs_se, h2_int	h2_int_se - or h2 Z and mean chi square which could be useful? Topher has a script that extracted h2 information for each GWAS

Select columns for supplementary tables
```{r create dataframe for supplementary tables}
rgs_supplmentary_table <- 
  rgs_supplmentary_table %>%
    select(
      Phenotype,
      rg, 
      se, 
      z,
      p,
      FDR_significant, ### add other methods too 
      Phenotype category,
      PMID, 
      code, 
      ancestry
      
```

4) Saving new tables
#what is the preffered format for saving the new table? google sheets? excel sheet for supplementary materials? Save summary table as rds for plotting?
Save as excel, google sheets, txt
```{r save supplementary table}

```

5) Plotting
Decide on significance threshold for plotting. Note: matrix decomposition threshold as less stringent.

Note: trait labels for plotting. This needs to be added to in the google spreadsheet, and then trait labels column for plotting will be used in stead of 'phenotype'

We also have multiple GWAS for one phenotype e.g. MDD - 
```{r vector of signficant traits for plotting}
# i.e. the researcher may only want to show specific GWAS for each phenotype e.g. most powerful, or certain cohorts, selected based on type of phenotyping. A
```


```{r filter based on vector of traits}

```


```{r create dataframe for plotting}
#filtering: 1) matrix decomposition 2) vector of selected traits for plotting
```

Use AN GWAS plot script (Watson et al 2019)
```{r plot rg results}

```

Publication table
```{r dataframe for publication table of significant rg traits}

```

save in excel format
```{r save publication table}

```
 
What output do we want from this script?
1) a general filtered summary table with full GWAS information for each trait (permissions, column significant)
2) an rg plot with top traits
3) a supplementary table with selected columns for paper (all traits, except traits with private permissions)
note: at some point we want to filter by heritability. But do we want this at an earlier step so that we are not running rg with traits that aren't significantly heritable?
4) Publication table - significant ones (instead of a plot)

