---
title: "rg results summary tables template"
author: "Abigail ter Kuile"
date: "28/08/2021"
output: html_document
---


#Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment=NA,
                      prompt=FALSE,
                      cache=FALSE)
```

Clear global environment
```{r Clear global environment}
remove(list = ls())
```

```{r Packages}
library(summarytools)
library(googlesheets4)
library(tidyverse)
```

Retrieve the recent date
```{r Recent date}
date = Sys.Date()
date
```

# Source credentials files
This file includes the file path to the King's College London Microsoft Teams channel to store the raw data files of the GWAS summary statistics (intermediate solution before we move the final version to the rosalind cluster). However, we probably will save all summary statistics raw files on SharePoint/Teams to save space on rosalind.
sumstats_path

The file also contains the login details for your google work account: google_account_details
```{r Credentials file}
source(file = "../credentials/credentials.R")
```

# Set output parameter for user

```{r}
opt<-list()

# This is an example output parameter
opt$output<-'/users/k_number/brc_scratch/Data/GWAS_sumstats/Rosalind_repository/example'

# Identify the directory for the output
opt$output_dir<-dirname(opt$output)

# Create output directory if it doesn't exist
if(!file.exists(opt$output_dir)){
  dir.create(opt$output_dir)
}
```

# GWAS summary statistics overview sheet

Login to googlesheets
```{r Login to google sheets}
gs4_auth(email = google_account_details)
```

Import google spreadsheet that contains the information of the GWAS
```{r Google spreadsheet of GWAS dictionary}
gwas_sheet <- read_sheet(
  ss = "https://docs.google.com/spreadsheets/d/1Wmj4Qzumkzks6541vQJqxaUBxjfL4KLQo7_K_oquacI/edit#gid=0",
  sheet = "SGDP_GWASLIST_EDITTHIS",
  col_names = TRUE,
  trim_ws = TRUE, # Trims whitespace,
  guess_max = 600,
  na = c("", "NULL", "NA")
  )
colnames(gwas_sheet)
gwas_sheet
```

#Input data for post-processing

What input do we need for this rmarkdown?
LDSC rg results between a selected trait of interest and all traits in the summary statistics repository (or a shorter list of specific list of traits). All traits have had mhc region removed before rg estimation (and are significantly heritable?). rg results are only calculated in one direction to avoid repetition. 


1) Read in list of traits to be analysed (in GWAS code format e.g. ADHD05, mhc region removed )
+++ATK this depends on how we decide we want to store out rg ldsc regression output results

Option 1: read in up to date list of all GWAS traits from rg analysis step

+++ATK This list will be created in a separate step, where traits are filtered for mean chi square 1.02 and mhc region removed
```{r read in up to date list of traits}
trait_list_all <- c(
  paste0(sumstats_path, "/data/full_trait_list.txt")
  ) 
```

Option 2: shorter list of specific traits for rg analysis that the user would add themselves. 
```{r read in sub selected trait list}
trait_list_subset <- c(
  paste0(file_path, "/data/subset_trait_list.txt")
  ) 
```

2) Read in rg data for your trait of interest

+++ATK this depends on how we decide to store out rg ldsc results. If we decide to store them as just log output files, then we can add bash script that we previously used here
```{r read in rg results}

```

```{r select rg results from log out put files}

```

2) Joining datasets:
#Step) join the rg results with the GWAS information file to get full GWAS info for each trait


+++ATK again, this is assuming we ar storing our rg output results separatly from our GWAS information sheet. 
+++ATK Ollie's script collage_gwas_info reads into a csv. Are we using csv or google sheet as the updated spread sheet?
```{r join info GWAS sheet with rg results}
rgs_gwas_info <- 
  left_join(
    x = rgs,
    y = GWAS_info_sheet,
    by = c("p1" = "Code")) ##note: p1 is for LDSC-python. Change if we use LDSC-Gsem to whatever we specify for that column (phenotype?)
    
colnames(rgs_gwas_info)
```

3) Filtering
For both summary table for plotting and supplementary table:

#step) Only extract GWAS traits that are not private. 
Filter by permissions and exclude all private sumstats

+++ATK: need to simplify the current permissions category 
```{r filter based on permissions}
rgs_filtered <- rgs_gwas_info %>% 
  filter(
    permissions != "PRIVATE" 
  )
```

#optional step if the user is only interested in certain trait categories
```{r by trait cateogry}
rgs_filtered_cat <- rgs_filtered %>%
  filter(
    category = "insert_category" #add category as relevant 
  )
```

#step) Filter p-values by bonferonni threshold 
Adapt Helena's script in python that already exists calculating bonferonni significance?
Create a new data frame with only bonf signif traits for summary table for paper plotting of top traits? or just add a new column with 'yes/no' for bonferonnni significant? Johan's preference: stick to one dataframe and filter in the plotting step
For supplementary table: how do we want to indicate which traits are bonferonni significant? e.g. bold text? A separate column saying 'significant/ non significant'? Colours??
```{r create bonferonni column}
rgs_filtered <- rgs_filtered %>%
  mutate(
   P.Bonferroni =
  p.adjust(
    p = p,
    method = "bonferroni" # Bonferroni correction method (p-values are multiplied by the number of comparisons)
    )
  )

#check
rgs_filtered %>% 
  summary(P.Bonferroni)
```

```{r create FDR column }
rgs_filtered <- rgs_filtered %>%
  mutate(
   FDR.Q.value =
  p.adjust(
    p = p,
    method = "BH" # Benjamini Hochberg method (the expected proportion of false discoveries amongst the rejected hypotheses)
    )
  )

#check
rgs_filtered %>% 
  summary(FDR.Q.value)
```

Add column with FDR significant label
```{r Add column with FDR significant label}
rgs_filtered <-
  rgs_filtered %>%
  mutate(
    FDR_significant =
      case_when(
        FDR.Q.value < 0.05 ~ "significant",
        FDR.Q.value > 0.05 ~ "not significant"
      )
  )

rgs_filtered %>%
  summarise(FDR_significant)
```

```{r convert table to matrix for matrix decomposition}
#find Helena's script for this 
```

note: we already have this script in python
```{r matrix decomposition for bonferonni threshold}
rgs_filtered <- rgs_filtered %>%
  mutate(
   P.Bonferroni_matrix = 
     )
     #find Helena's script for this?
```

Create new dataframes for supplmentary tables

#create supplementary tables
Select columns for supplementary tables for paper. Decide on columns. 
e.g. Watson et al 2019 (AN GWAS) columns: Phenotype, rg, se, P (green = bonferoni significant), Phenotype category, PMID, code, ancestry. 
but could also include Z value of rg, gcov_int, gcov_int_se, h2_obs, h2_obs_se, h2_int	h2_int_se - or h2 Z and mean chi square which could be useful? Topher has a script that extracted h2 information for each GWAS

Select columns for supplementary tables
+++ATK update column renaming when we know what input we are reading results from (GWAS info sheet and rg results)
```{r create dataframe for supplementary tables}
rgs_supplmentary_table <- 
  rgs_filtered %>%
    select(
      Phenotype,
      PMID,
      mean_chi^2,
      h2_obs, #heritability information for each trait
      h2_obs_se,
      h2_int,
      h2_int_se,
      h2_z, 
      rg, #genetic correlations result for each trait with trait of interest
      rg_se, 
      rg_z,
      p,
      P.Bonferroni,
      FDR_significant,
      P.Bonferroni_matrix,
      Phenotype_category,
      code, #trait code
      ancestry
      ) # +++ATK: do we want ancestry?
      
#Check
rgs_supplmentary_table %>%
  glimpse()
```

4) Saving new tables
#what is the preffered format for saving the new table? google sheets? excel sheet for supplementary materials? Save summary table as rds for plotting?
Save as excel, google sheets, txt
```{r save supplementary table as csv}
# Save new/updated gwas_info
rgs_supplmentary_table %>% 
  write_csv(
    file = paste0(opt$output,
                  "rgs_table_",
                  Sys.Date(),
                  ".csv")
  )

```

```{r save supplementary table as rds}

rgs_supplmentary_table %>% 
  saveRDS(
    file = paste0(opt$output,
                  "rgs_table_",
                  Sys.Date(),
                  ".rds")
    )
```



5) Plotting

Notes for the user on deciding which traits to include for plotting:
1) If only showing significant traits, decide on significance threshold. Note: matrix decomposition threshold is less stringent.
2) If more than one GWAS is available for a phenotype, the researcher may only want to show specific GWAS for each phenotype e.g. most powerful, or certain cohorts, selected based on type of phenotyping. 

Note: trait labels for plotting. This needs to be added to in the google spreadsheet, and then trait labels column for plotting will be used in stead of 'phenotype'
```{r Check names of significant traits}
rgs_supplmentary_table %>% 
  filter
```


```{r vector of signficant traits for plotting}
traits_plotting <- c("trait name 1",
                     "trait name 2"
                     )
```


```{r filter based on vector of traits}
dat %>% 
  select(all_of(traits_plotting))
```


```{r create dataframe for plotting}
#filtering: 1) matrix decomposition 2) vector of selected traits for plotting
```

Use AN GWAS plot script (Watson et al 2019)
```{r plot rg results}
plot <- ggplot(data = dat, aes(x=(dat$Phenotype2), 
                                 y=dat$rg, 
                                 label=rg)) +
                 theme_linedraw() + #change to theme_bw if want to save as pdf if linedraw theme wont save as pdf
                 geom_bar(stat='identity', aes(fill=Category), width=.5) +
  geom_errorbar(aes(ymin =  rg-se, 
                    ymax = rg+se),
                width = 0.2,
                size=0.3,
                colour="#303535") + 
  scale_fill_manual(values = coul) +
labs(subtitle="Ordered by genetic correlation with (insert trait) within each category", 
       title= "Genetic correlations between (insert trait) and heritable traits",
       x="Other phenotype\n",
       y="Genetic correlation") +
  theme(plot.title = element_text(size=12,
                                  hjust = 0,
                                  face="bold"),
        plot.subtitle = element_text(size=12,
                                          hjust = 0,
                                face="italic"),
        axis.text = element_text(size=10),
         plot.caption = element_text(hjust = 0,
                                          size =12,
                                         face="italic"),
        axis.title.y=element_blank(),
        legend.title = element_text(face='bold'),
        axis.line = element_line(size=1))                          
  scale_y_continuous(limits=c(-1,1),
                     breaks=seq(-1,1,0.2))                          
    plot + coord_flip()
    ggsave(filename = "rgs_plot_.png")
```

Publication table
```{r dataframe for publication table of significant rg traits}

```

save in excel format
```{r save publication table}

```
 
What output do we want from this script?
1) a general filtered summary table with full GWAS information for each trait (permissions, column significant)
2) an rg plot with top traits
3) a supplementary table with selected columns for paper (all traits, except traits with private permissions)
note: at some point we want to filter by heritability. But do we want this at an earlier step so that we are not running rg with traits that aren't significantly heritable?
4) Publication table - significant ones (instead of a plot)

